---
title: "Homework 2: k Sample Inference"
body-classes: hw2-page
author: "Lions like to eat onions"
format:
  html:
    number-sections: true
    code-fold: true
    code-line-numbers: true
    cap-location: top

  pdf:
    documentclass: article
    number-sections: true
    reference-section-title: "References"
    code-line-numbers: true
    cap-location: top
    geometry:
      - top=0.75in
      - left=0.75in
      - bottom=0.75in
      - right=0.75in

format-links: [pdf]
---

## Question 1

### Part a
#### 1
```{r}
set.seed(354)

R <- 1000
simulations <- vector("list", R)

for (i in 1:R) {
  simulations[[i]] <- data.frame(
    value = rnorm(150, mean = 0, sd = 1),
    group = factor(rep(1:5, each = 30))
  )
}
```

#### 2
```{r}
error_unadjusted <- logical(R)

for (i in 1:R) {
  dat <- simulations[[i]]

  test_unadjusted <- pairwise.t.test(
    dat$value, dat$group,
    p.adjust.method = "none"
  )

  error_unadjusted[i] <- any(
    test_unadjusted$p.value < 0.05,
    na.rm = TRUE
  )
}

mean(error_unadjusted)
```

At least one pairwise comparison provided statistically discernible
support for a non-zero mean difference in
`r sum(error_unadjusted)` of the 1,000 simulations (p < 0.05).
These are Type I errors,
giving an estimated family-wise error rate of
`r round(mean(error_unadjusted) * 100, 1)`%.

#### 3

```{r}
error_tukey <- logical(R)

for (i in 1:R) {
  dat <- simulations[[i]]

  fit <- aov(value ~ group, data = dat)
  test_tukey <- rstatix::tukey_hsd(fit)

  error_tukey[i] <- any(test_tukey$p.adj < 0.05)
}

mean(error_tukey)
```

After Tukey's HSD adjustment, at least one pairwise comparison
provided statistically discernible support for a non-zero mean
difference in `r sum(error_tukey)` of the 1,000 simulations
(adjusted p < 0.05).
Since all population means were zero, these are Type I errors,
and the estimated family-wise error rate is
`r round(mean(error_tukey) * 100, 1)`%.


### Part b

```{r}
error_bh <- logical(R)

for (i in 1:R) {
  dat <- simulations[[i]]

  test_bh <- pairwise.t.test(
    dat$value, dat$group,
    p.adjust.method = "BH"
  )

  error_bh[i] <- any(
    test_bh$p.value < 0.05,
    na.rm = TRUE
  )
}

mean(error_bh)
```

After BH adjustment, at least one pairwise comparison provided
statistically discernible support for a non-zero mean difference
in `r sum(error_bh)` of the 1,000 simulations, giving an estimated
family-wise error rate of `r round(mean(error_bh) * 100, 1)`%.
This was much lower than the unadjusted rate of
`r round(mean(error_unadjusted) * 100, 1)`% and close to Tukey's
rate of `r round(mean(error_tukey) * 100, 1)`%.

BH targets the false discovery rate (FDR), whereas Tukey's HSD
controls the family-wise error rate (FWER). 

## Question 2

### Part a

```{r}
# Required sample size per group
pwr::pwr.anova.test(
  k = 5,
  f = 0.25,
  sig.level = 0.05,
  power = 0.80
)

# Power curve
n <- 2:100

power <- sapply(n, function(x) {
  pwr::pwr.anova.test(
    k = 5, n = x, f = 0.25, sig.level = 0.05
  )$power
})

plot(n, power, type = "l",
     xlab = "Sample size per group",
     ylab = "Power",
     main = "ANOVA Power Curve")

abline(h = 0.80, col = "red", lty = 2)
abline(v = 40, col = "blue", lty = 2)
```

The required sample size is roughly 40 participants per group and 200 in total.
The graph shows that 40 participants per group achieves slightly
more than 80% power.

Since every group in the original groups had more than 40
participants, the study is powered for the overall
ANOVA under the assumed moderate effect size. This calculation
assumes equal group sizes, so it does not provide the exact power
of the original unequal-sized design.


### Part b

```{r}
# Non-centrality parameter and power
lambda <- 150 * 0.25^2
lambda

critical <- qf(0.95, 4, 145)
pf(critical, 4, 145, ncp = lambda, lower.tail = FALSE)

# Alternative F distribution
curve(df(x, 4, 145, ncp = lambda), from = 0, to = 10,
      xlab = "F", ylab = "Density",
      main = "Alternative F Distribution: N = 150")
abline(v = critical, col = "red", lty = 2)
```

With 30 participants per group, the non-centrality parameter is
9.375 and the power is approximately 66.8%. Therefore, N = 150
is insufficient to achieve the desired 80% power.

### Part c

```{r}
# Power for different sample sizes per group
n <- 40:500

power_c <- sapply(n, function(x) {
  PMCMRplus::power.tukey.test(
    n = x,
    groups = 5,
    delta = 0.25,
    within.var = 0.7152,
    sig.level = 0.05
  )$power
})

# Power with 40 participants per group
power_c[1]

# Smallest sample size achieving at least 80% power
n_required <- min(n[power_c >= 0.80])
n_required
```

An overall ANOVA can detect differences among groups having different
means. However, detecting the smaller difference between
stages 1 and 2 is harder, especially after Tukey's adjustment
for multiple comparisons.

With 40 participants per group, the power is only about 7.9%.
Using the pooled variance of 0.7152, approximately 293 participants
per group are needed to achieve 80% power for this comparison.

### Part d

```{r}
#| cache: true
#| warning: false
library(Superpower)

for (n in c(40, 80)) {

  design <- ANOVA_design(
    design = "5b",
    n = n,
    mu = c(2.92, 3.11, 3.36, 4.04, 4.35),
    sd = sqrt(0.80),
    plot = FALSE
  )

  result <- ANOVA_power(
    design,
    nsims = 1000,
    seed = 354,
    emm = TRUE,
    emm_p_adjust = "tukey",
    verbose = FALSE
  )

  print(paste("Sample size per group:", n))
  print(result$main_result)
  print(result$emm_results)
}
```

Doubling the sample size from 40 to 80 per group increases power,
but is still insufficient to achieve 80% power for every pairwise
comparison. In specific, the small difference between stages
0 and 1 (3.11 − 2.92 = 0.19) remains difficult to detect.